`{peacesciencer}`: An R Package for Quantitative Peace Science Research

### 2021-11-11 user-guide/ms.Rmd
update site

This data feature tackles a recurring problem for researchers in the peace science community. True research reproducibility is best achieved creating data from scratch, though no published guide exists that informs researchers how to do this on their own. Instead, researchers may end up reusing old code generated in past studies, leaving them to spend time and energy adjusting the sample of states, the temporal domain, and doing whatever additional troubleshooting may arise from this practice. Researchers may additionally spend too much time reproducing old code for standard information that goes into any dyadic or monadic analysis---like contiguity relationships and democracy---and have to do additional troubleshooting for how these various data sources treat missing data or treat state codes in a manner inconsistent with the more accessible Correlates of War or Gleditsch-Ward state codes. This is all compounded by changes in technology that treat the creation of the data and the analysis of data as a continuous process in which the contemporary quantitative political scientist is increasingly becoming a computer programmer as well. Graduate students, in particular, face unique challenges associated with these developments. Graduate students beginning their careers in peace science must learn how scholarship informs data in peace science and how data inform scholarship at the same time they are needing to learn quantitative methods in a chosen software package.

`{peacesciencer}` is designed to address these problems. Built around the free and open source R programming language, `{peacesciencer}` contains a suite of data and functions for creating data of interest to researchers. Researchers can use `{peacesciencer}` to create dyad-year, leader-year, leader-dyad-year, and state-year data (among some others) from scratch. Afterward, they can add a variety of standard information (e.g. contiguity, alliances, major power status, GDP per capita estimates, capability estimates, and more) to these data with a simple command. This is a considerable time-saver since, in the absence of it, researchers would have to more meticulously code and transform the raw data to conform to the kind of data they want. `{peacesciencer}` comes with some data innovations as well, including a comprehensive data set on democracy by year, an original data set on capitals and capital transitions, and a function to create peace years between ongoing conflicts. All are done with the maximum possible transparency. The project is available for public view of Github (`r params$ps_github`). The `data-raw` directory on the project's Github contains information and comments about how every data set was created and the `R` directory informs the user how each function works. The function manuals (`r params$ps_manual`) contain additional comments about what each function returns and, in appropriate cases, why it is doing what it is doing. Thus, `{peacesciencer}` not only assists a peace scientist with their research, but it does so in a manner that best conforms to the  Data Access and Research Transparency Initiative (DA-RT) initiative across all political science.

This data feature proceeds in the following fashion. The next section expands what need this package fills for peace scientists in the present. Afterward, it provides an overview of what is included in `{peacesciencer}` to help researchers more quickly conduct the kind of quantitative research they want. Thereafter, it provides a tutorial on how to install and best use `{peacesciencer}` in the R programming language. A more comprehensive tutorial follows, showing how `{peacesciencer}` already has a suite of data and functions that can allow for effective replications of a "dangerous dyads" type analysis [@bremer1992ddc], standard state-year analyses of civil conflict onset [e.g. @fearonlaitin2003eicw], and even leader-year analyses of inter-state dispute initiation [@horowitzstam2014hpme]. This feature concludes with a comparison of `{peacesciencer}` with other alternatives and a discussion of how `{peacesciencer}` can inform more reasoned design decisions for researchers in peace science.

@@

`{peacesciencer}` is motivated by the following ideals that led to its creation. For one, researchers invest too much time in the construction of a data set that faithfully captures the unit of analysis. Assume a researcher wants an original data set on all directed dyad-years for Correlates of War states for an analysis of inter-state conflict. How might one do that? The answer has never been immediately obvious. No published guide exists that shows a researcher how to create these data themselves from scratch, which is one reason why software bundles like EUGene and NewGene are attractive to researchers who primarily care about the substance of their research question. After all, EUGene's main value---if not its primary impetus---was allowing researchers to create data sets for replication of previous studies, producing a host of data types (e.g. dyad-year, state-year, dispute-year) along the way that users can amend as they saw fit. `{peacesciencer}` is primarily born from this question about how to create these data from scratch. The underlying code that produces these data types is available online and `{peacesciencer}` converts these lines of code into simple functions for the ease of the researcher.

@@

Third, the creation of the data and the analysis of the data are increasingly becoming one continuous process. Not too long ago, it used to be the case that researchers had to download a data set, or create one from scratch (possibly in a spreadsheet or through a program like EUGene). After downloading or constructing the data, the researcher then opened a specialty program for statistical analysis (e.g. SAS, SPSS, Stata) to recode raw data into a form suitable for analysis before running a statistical model that regresses some outcome on a set of covariates. Current research practices still resemble this process, but the steps between them are no longer as large as they were in the past. Software options exist that allow the researcher to load data, create data, clean data, analyze data, and present the results of the analysis all within one program. `{peacesciencer}`, by itself, does not do all these things, but it seamlessly connects the beginning of the research process to the end of the research process without needing to leave the increasingly popular R programming language and Rstudio (its free-for-use integrated development environment [IDE]).

@@

Table \@ref(tab:tab-createfunctions) are functions that create base data frames for a researcher starting an original project. They serve as functions that communicate the units of analysis supported in this package and that the package is capable of generating for an interested user. For example, `create_stateyears()` will generate the full universe of state-years from the Correlates of War [@cowstates2016, v. 2016] or Gleditsch-Ward [@gledtischward1999rlis, v.2017] system, encompassing all state-years to the most recently concluded calendar year, depending on the arguments supplied to the user in the function. `create_leaderyears()` will generate the full universe of leader-years from the Archigos leader data [@goemansetal2009ia, v. 4.1], optionally standardizing leader-years to the Gleditsch-Ward or Correlates of War state system data. As of version 0.7, `{peacesciencer}` is capable of creating full dyad-year data, leader-day data, leader-dyad-year data, leader-year data, state-day data, and state-year data.[^leaderdyadyear] A vignette on the package's website shows how users can create other forms of data from these functions as well (e.g. leader-months, state-quarters).

@@

`{peacesciencer}`'s functions work out of the box, though users should find their experience augmented by two additional downloads. First, RStudio offers an IDE that serves as a user-friendly graphical user interface (GUI) over what is, at its core, a programming language with a command-line interface. Rstudio's design will make it much easier for users to experiment with `{peacesciencer}`'s functionality and read its documentation to assist them with the user of these functions. The second additional download is `{tidyverse}`, itself a suit of packages that share a common form and design [@wickhametal2019wt]. `{peacesciencer}` functions make considerable use of the component packages of `{tidyverse}` but installing and loading `{tidyverse}` will allow the researcher to make quicker use of `{peacesciencer}`'s functionality. A user can open an R session by way of Rstudio and install both packages as follows.

@@

I encourage users who are using `{peacesciencer}` for the first time, especially those who are learning R for the first time because of their interest in this package, to approach `{peacesciencer}` with an idea of the kind of data they want to create for the sake of a project. The core functionality of `{peacesciencer}` begins with the creation of a data frame, after which can be populated with various indicators of interest. No matter, the suggested use of `{peacesciencer}` begins with the creation of a data frame. To start, assume a new user without much familiarity with the R programming language installed Rstudio, `{tidyverse}`, and `{peacesciencer}` with the idea of using `{peacesciencer}` to help them start a new research project that seeks to explain civil conflict onset across state-years. Toward that end, they have identified their unit of analysis is state-year and can be created with the `create_stateyears()`. To get started, I encourage entering the following command in the console in Rstudio.

>>>>

This data feature tackles a recurring problem for researchers in the peace science community. True research reproducibility is best achieved creating data from scratch, though no published guide exists that informs researchers how to do this on their own. Instead, researchers may end up reusing old code generated in past studies, leaving them to spend time and energy adjusting the sample of states, the temporal domain, and doing whatever additional troubleshooting may arise from this practice. Researchers may additionally spend too much time reproducing old code for standard information that goes into any dyadic or monadic analysis---like contiguity relationships and democracy---and have to do additional troubleshooting for how these various data sources treat missing data or treat state codes in a manner inconsistent with the more accessible Correlates of War or Gleditsch-Ward state codes. This is all compounded by changes in technology that treat the creation of the data and the analysis of data as a continuous process in which the contemporary quantitative political scientist is increasingly becoming a computer programmer as well. Graduate students and other beginners in the field face unique challenges associated with these developments. Students just learning peace science must learn how scholarship informs data in peace science and how data inform scholarship at the same time they are needing to learn quantitative methods in a chosen software package.

`{peacesciencer}` is designed to address these problems. Built around the free and open source R programming language, `{peacesciencer}` contains a suite of data and functions for creating data of interest to researchers. Researchers can use `{peacesciencer}` to create dyad-year, leader-year, leader-dyad-year, and state-year data (among some others) from scratch. Afterward, they can add a variety of standard information (e.g. contiguity, alliances, major power status, GDP per capita estimates, capability estimates, and more) to these data with a simple command. This is a considerable time-saver since, in the absence of it, researchers would have to more meticulously code and transform the raw data to conform to the kind of data they want. `{peacesciencer}` comes with some data innovations as well, including a comprehensive data set on democracy by year, an original data set on capitals and capital transitions, and a function to create peace years between ongoing conflicts. All are done with the maximum possible transparency. The project is available for public view on Github (`r params$ps_github`). The `data-raw` directory on the project's Github contains information and comments about how every data set was created. The function manuals (`r params$ps_manual`) contain additional comments about what each function returns and, in appropriate cases, why it is doing what it is doing. Thus, `{peacesciencer}` not only assists a peace scientist with their research, but it does so in a manner that best conforms to the Data Access and Research Transparency Initiative (DA-RT) initiative across all political science.

This data feature proceeds in the following fashion. The next section expands what need this package fills for peace scientists. Afterward, it provides an overview of what is included in `{peacesciencer}` to help researchers more quickly conduct the kind of quantitative research they want. Thereafter, it provides a tutorial on how to install and best use `{peacesciencer}` in the R programming language. A more comprehensive tutorial follows, showing how `{peacesciencer}` already has a suite of data and functions that can allow for effective replications of a "dangerous dyads" type analysis [@bremer1992ddc], standard state-year analyses of civil conflict onset [e.g. @fearonlaitin2003eicw], and even leader-year analyses of inter-state dispute initiation [e.g. @horowitzstam2014hpme]. This feature concludes with a comparison of `{peacesciencer}` with other alternatives and a discussion of how `{peacesciencer}` can inform more reasoned design decisions for researchers in peace science.

@@

`{peacesciencer}` is motivated by the following observations and ideals that led to its creation. For one, researchers invest too much time in the construction of a data set that faithfully captures the unit of analysis. Assume a researcher wants an original data set on all directed dyad-years for Correlates of War states for an analysis of inter-state conflict. How might one do that? The answer has never been immediately obvious. No published guide exists that shows a researcher how to create these data themselves from scratch, which is one reason why software bundles like EUGene and NewGene are attractive to researchers who primarily care about the substance of their research question. After all, EUGene's main value---if not its primary impetus---was allowing researchers to create data sets for replication of previous studies, producing a host of data types (e.g. dyad-year, state-year, dispute-year) along the way that users can amend as they saw fit. `{peacesciencer}` is primarily born from this question about how to create these data from scratch. The underlying code that produces these data types is available online and `{peacesciencer}` converts these lines of code into simple functions for the ease of the researcher.

@@

Third, the creation of the data and the analysis of the data are increasingly becoming one continuous process. Not too long ago, it used to be the case that researchers had to download a data set, or create one from scratch (possibly in a spreadsheet or through a program like EUGene). After downloading or constructing the data, the researcher then opened a specialty program for statistical analysis (e.g. SAS, SPSS, Stata) to recode raw data into a form suitable for analysis before running a statistical model that regresses some outcome on a set of covariates. Current research practices still resemble this process, but the steps between them are no longer as large as they were in the past. Software options exist that allow the researcher to load data, create data, clean data, analyze data, and present the results of the analysis all within one program. `{peacesciencer}`, by itself, does not do all these things, but it seamlessly connects the beginning of the research process to the end of the research process without needing to leave the increasingly popular R programming language and RStudio (its free-for-use integrated development environment [IDE]).

@@

Table \@ref(tab:tab-createfunctions) are functions that create base data frames for a researcher starting an original project. They serve as functions that communicate the units of analysis supported in this package and that the package is capable of generating for an interested user. For example, `create_stateyears()` will generate the full universe of state-years from the Correlates of War [@cowstates2016, v. 2016] or Gleditsch-Ward [@gledtischward1999rlis, v.2017] system, encompassing all state-years to the most recently concluded calendar year, depending on the arguments supplied to the user in the function. `create_leaderyears()` will generate the full universe of leader-years from the Archigos leader data [@goemansetal2009ia, v. 4.1], optionally standardizing leader-years to the Gleditsch-Ward or Correlates of War state system data. As of version 0.7, `{peacesciencer}` is capable of creating full dyad-year data, leader-day data, leader-dyad-year data, leader-year data, state-day data, and state-year data.[^leaderdyadyear] A vignette on the package's website shows how users can create other forms of data from these functions as well (e.g. dyadic-dispute-year, leader-months, state-quarters).

@@

`{peacesciencer}`'s functions work out of the box, though users should find their experience augmented by two additional downloads. First, RStudio offers an IDE that serves as a user-friendly graphical user interface (GUI) over what is, at its core, a programming language with a command-line interface. RStudio's design will make it much easier for users to experiment with `{peacesciencer}`'s functionality and read its documentation to assist them with the use of these functions. The second additional download is `{tidyverse}`, itself a suit of packages that share a common form and design [@wickhametal2019wt]. `{peacesciencer}` functions make considerable use of the component packages of `{tidyverse}` but installing and loading `{tidyverse}` will allow the researcher to make quicker use of `{peacesciencer}`'s functionality. A user can open an R session by way of RStudio and install both packages as follows.

@@

I encourage users who are using `{peacesciencer}` for the first time, especially those who are learning R for the first time because of their interest in this package, to approach `{peacesciencer}` with an idea of the kind of data they want to create for the sake of a project. The core functionality of `{peacesciencer}` begins with the creation of a data frame, after which can be populated with various indicators of interest. No matter, the suggested use of `{peacesciencer}` begins with the creation of a data frame. To start, assume a new user without much familiarity with the R programming language installed RStudio, `{tidyverse}`, and `{peacesciencer}` with the idea of using `{peacesciencer}` to help them start a new research project that seeks to explain civil conflict onset across state-years. Toward that end, they have identified their unit of analysis is state-year and can be created with the `create_stateyears()`. To get started, I encourage entering the following command in the console in RStudio.
